Query Optimization for Dynamic Imputation

### 2017-05-11 rebuttal.tex

\documentclass{vldb}
\usepackage{xcolor}
\newcommand{\resp}[1]{{\color{blue}{#1}}}
\newcommand{\todo}[1]{{\color{red}{\textit{#1}}}}
\newcommand{\ProjectName}[0]{ImputeDB}

\begin{document}
\section{Multiple reviewers}
\begin{itemize}
\item Redundant figures, excessive whitespace, formatting.

  \resp{We've compressed many of the figures and removed whitespace to make room for additional material.}
\end{itemize}

  \section{Reviewer 1}
  \begin{itemize}
  \item W2. The paper completely obscures the ML technique being used: (1) It would be nice to have some discussion of the complete API for implementing an ML technique, as well as how one might implement a cost-estimator for one or two examples. (2) It would be helpful to see the impact of a few different approaches (e.g., Tree-based, Interpolation, Regression, FD-based) on performance and accuracy. (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.
    \todo{}

  \item W3. Reducing runtime to ~10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE). Worse still, there's no way for the user to differentiate between the cases of Q4 and Q8 at runtime. 24\% error relative to direct imputation is huge.

    \resp{
      The outsized error in Q8 is a result of a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
      groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
      using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.
      \todo{We have extended ImputeDB to provide users with a estimate of the number of tuples dropped in a calculation due to nulls, which may be indicative of the underlying error or skew
        in the results presented. This is a simple heuristic but we find that it would have provided appropriate warning in cases such as Q8.}
    }

  \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
    \todo{}

  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach — between 0 and 24 percent" 24 percent error is pretty bad.
    \todo{}

  \item S2: The drop operator appears to be a weaker form of SQL's null-value / 3-valued Boolean semantics. The latter might actually be more useful. For example, consider the case of a predicate (A $>$ 5 OR B $<$ 10). If A $>$ 5, then the condition is independent of B, and the tuple can still be salvaged. If A $\leq$ 5, then the expression evaluates to UNKNOWN and the tuple is dropped anyway. For aggregates there is a greater need, as one might have different aggregates computed over different subsets of the data, but even here it's not clear that this is any less informative than just throwing data away.

@@

    \resp{We agree that it could be interesting to allow drops to be omitted before filters. This would increase the plan space somewhat and might lead to better plans in some cases. For the conjunctive predicates in our experiments, our implementation splits the predicate into multiple filter nodes, so imputations or drops and filters can interleave. This should have the same effect, as the filter and the drop operator behave the same way on a tuple containing null. However, this optimization would also apply to disjunctive predicates, which cannot be split, and is worth definitely investigating.}

  \item S2: Why no unions?

    \resp{We skipped unions for brevity. There is nothing fundamental about our algorithm which makes unions difficult to implement.}

  \item S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss. It would help to clarify that Loss is not about tuple loss, but rather a heuristic penalty function.

    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to \textit{Penalty}.}

  \item S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data. I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set. For example, consider a one-to-many join \verb/µ(R) |X| S/. Your optimization algorithm would likely wish to move imputation up: \verb/µ(R |X| S)/. Yes, there is more training data, but if the fanout for each tuple is skewed, some tuples in R will become over-represented in the join result. A similar observation holds for selection. For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree $\geq$ 0" presumably indicate invalid data that should not be used for training. (is this perhaps why you have so much error for full imputation on Q8?)

@@

    \resp{This is a reasonable concern. Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query requires the construction a large, skewed data set. The problem seems to be particular to queries which generate large intermediate tables. In some cases, as when joining two datasets, the intermediate table is a good target for training because it exposes correlations which would be hidden in the separate tables. In other cases, this intermediate table is large and skewed, which causes problems for imputation. Avoiding this case requires us to estimate the skew in these intermediate tables.

      To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$. This was in fact a trivially satisfied predicate given the way our dataset was set up.}
    \todo{Need a better explanation here.}

  \item S3.4.4: It appears that Time() assumes that plans are IO-bound. This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.

@@

    \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function. We have clarified this in the paper.}

  \item S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator. Why not just take the same approach as selection and treat it as having negligible cost?
    \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
      our implementation, it has $0$ cost, and we have updated the text to reflect that.}

  \item S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?

    \resp{Though we do not show this explicitly, we repeat all of the queries using COUNT
      aggregates in order to compute the ``Count Fraction''. The timing results are
      clearly identical in this case and the number of tuples returned can be seen
      directly in Table 4.}

    \todo{For SUM queries, that may depend more directly on the number of rows returned than the
      quality of imputations for any individual row, the effect of different COUNT results
      dominates the computation AVG*COUNT. How do we handle this?}

  \item S4.4.1: Imputation over the raw data is being used as ground truth. Is there any reason to suspect that this is the correct thing to do? Perhaps joining multiple tables together exposes new attributes that makes imputation *more* reliable? At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.

@@

    \resp{For our purposes, imputation over the raw data is a reasonable ground truth. This
      is because the analyst, in practice, should be considering the tradeoffs on quality
      and performance of using ImputeDB versus a production imputation system on the raw
      data. Unfortunately, they would never have access to a real ground truth in such a
      scenario, and moreover, finding ground truth for use even in an empirical evaluation
      of imputation methods can be challenging because artificially dirtying the data
      often does not capture the complex reasons for missingness in real world datasets.
      We update the discussion in S4.4.1 to emphasize this point.}

  \item There are a number of opportunities for trimming in the figures, tables, and algorithms: Algorithms 1-3 could be compacted a bunch, Table 1 is full of whitespace (e.g., make it a 3-column {figure*}), Tables 1-2 could also be summarized as CDFs, Figure 6 is full of whitespace, Table 3 could be put queries on a single row (e.g., Q1 would easily fit on 1 line), and the references have a lot of redundant information (URLs, for example).

@@

    \resp{We have compacted the algorithms and tables. Note that Tables 1-2 show the percentage of missing values for \emph{each attribute}, not as an overall percentage, so cannot be summarized as CDFs. We think that the URLs add value to the references, particularly for data sets.}

  \item Additional space recovered from these sections could be used to discuss the choice of techniques for imputation, and the confidence estimation idea suggested in the future work section.

  \item Another interesting use of the space, or at least an idea for future work might be the role of aggregation on inference accuracy. In my experience aggregate queries tend to be *much* more forgiving of approximations (e.g., Online Aggregation) and data errors.
  \end{itemize}

@@

  \section{Reviewer 2}
  \begin{itemize}
  \item 1) No new imputation method is presented. The model relies on external imputation techniques, and given it's done on-the-fly, presumably imputed values depend on qualifying query records.

    \resp{Our contribution is the planning algorithm which allows imputation to be interleaved into the query plan.
      We do not present a new imputation method, but our technique is agnostic to the particular imputation method used.
      We reduce the overhead of imputation by performing it only on the tuples which are relevant to a particular query, so the imputed results do depend on the records which are relevant to particular query.}

@@

  \item 2) No comparative evaluation against existing imputation/data cleaning techniques.

    \resp{
      We compare our planning technique for on-the-fly imputation to imputation performed on the full dataset, using the same imputation algorithm in both cases.
      We are not aware of another database technology which offers on-the-fly imputation, so there is no direct comparison for the planning algorithm.
      % Our evaluation compares our technique to an imputation of the full dataset. We use the same imputation algorithm for our on-the-fly imputation as for the full dataset imputation. Cleaning the full dataset is the standard method of handling missing values.
    }

@@

  \item D1) The model relies on external imputation techniques, and given it's done on-the-fly, the imputed values will differ based on the qualifying query records?

    \resp{Yes, as discussed in the response to 1.}

  \item D2) The paper does not propose new imputation techniques but rely on existing statistical imputation algorithms. There is no comparative evaluation, nor samples of imputed values to show the goodness of the imputed values. It would be nice to see a comparative evaluation of different statistical imputation methods, and suggestions for extensions that can fit within the proposed imputation model.

    \resp{
      We have added two new imputation methods to the experiments, to show that the planning algorithm can be used successfully with different imputation methods.
      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving rules for choosing cost functions for a new imputation method.
    }
  \end{itemize}

  \section{Reviewer 3}
  \begin{itemize}
  \item W2. More work is required in the presentation: important parts are not explained in depth, while there are many details in other sections that can be omitted.

@@

    \resp{Yes.}

  \item D6. During plan enumeration, can you safely prune plans at some point, similar to what a traditional optimizer would do?

    \resp{We maintain the set of pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning. We found that the size of these pareto sets is small enough that we did not need to prune further.

      If necessary, it is possible to sample from the pareto optimal plans during the plan enumeration. The final selected plan may not be pareto optimal, but this is a potential way to control the number of plans.}
    \todo{Talk about pareto set approximation here.}

  \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?
    \todo{Micah: what do we actually do here?}

@@

  \item D9. Additional experiments I suggest to include:
    -- Consider a scenario where the user submits multiple similar queries. Show after how many such queries your approach has no longer a competitive advantage compared to performing imputation on the data.
    -- Try different values for $\alpha$ (not just 0 and 1).
    -- Show runtime of enumeration algorithm as queries get more complex.

    \resp{
      The sensitivity of different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
      a lot of the same plans for execution. In lieu of this, we extended the results section to include a subsection that presents the final frontiers for
      two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.
      \\
      We explored the planning times for queries with increasing number of joins. The standard algorithm presented in the original version submitted handles
      up to 5 joins, with columns requiring imputation in all tables, with planning time of approximately one second. We added a small optimization to help reduce
      planning times for a larger number of joins. We maintain an approximation of the Pareto sets, rather than the exact set, by only adding non-dominated plans that
      are sufficiently different. This helps reduce some of the increases in planning time for a larger number of joins.}

  \item D10. The scenario you are considering, with multiple queries during exploration, begs for reuse of previous query results, as you also mention in the future work section.

    \resp{We agree.
      Reusing previous query results is a tricky problem because it requires the database to speculatively store some of the intermediate results of its query plans.
      A simple heuristic which should work well is to store the results of full table imputation whenever a query forces this to occur. These results are guaranteed to be useful later.
      Storing other intermediate results is problematic, particularly those which are computed after filtering or joining, as we would need to have a mapping back to the tuples in the base tables.}

@@

  \item D11. Less important points:
    -- Sections 4.2 and 4.3 in the experiments are too verbose.
    -- In Section 3.2, you mention that joins are performed after filtering. When filters are on results of joins, this is not possible.

    \resp{The subset of SQL that we consider does not allow filtering on the results of joins (see Section 3.2). Extending the planner to cover this case is not complex. Before adding filters, it must add imputations on the required attributes.}

@@

  \item -- Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.

    \resp{Yes, Tummer \& Koch is definitely related.
      However, much of the complexity of their method is in dealing with cost functions which are not well behaved.
      Our cost functions are monotonic, which allows us to use a simpler optimizer than what Tummer \& Koch propose.
    We have added a discussion of their work to the related works section.}

### 2017-05-11 paper/algorithm.tex
Update formulae
paper: continue to reduce figures
a pass over cost model and impute api

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
$\textsc{Penalty}(q) \in (0,1]$ is a heuristic that attempts to quantify the amount of information lost by the imputation procedure used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics, selectivity estimation of the query predicates, and the complexity of the
impute operation.

@@

This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all possible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.

@@

Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$.

So in essence, $\alpha$ is a tunable parameter that determines whether the optimizer focuses on quality or on performance, as plans that
lose a lot of information through dropping are also the fastest.

@@

\noindent{\textbf{Constructing $P$}}: The form of $P$ is chosen to reflect the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations to summary statistics such as mean and variance
Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
The heuristic penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases. Finally, the heuristic penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.

@@

\noindent{\textbf{Constructing $T$}}: As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on
the properties of the underlying algorithm via a function $T$. If we assume
that the \textit{Impute} routine is CPU-bound, then we will adjust
\textsc{Time} by some constant factor $f$, thereby scaling the result to be
comparable with IO-bound routines.

The form of $T$ depends on the specific imputation method, and as with $P$,
there are some principles to capture the intuition.  $T$ should \emph{increase}
as the number of tuples and complete and dirty attributes passed to the
imputation function increases and it should be take into account the complexity
of the imputation algorithm. Analysis of the time complexity of different
imputation algorithms, especially with in-database execution, is beyond the
scope of this paper. As with $P$, it is the ordering between queries
generated by $T$ that is more important than the precision of estimates.

@@

\subsection{Imputation strategies}
\label{sec:imputation}

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
penalty and time functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These can then be queried by the optimizer during planning.  The flexibility of
this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation.

@@

New imputation algorithms can also be added. As a reference implementation,
\ProjectName{} uses a general-purpose imputation strategy based on
chained-equation decision trees.  Chained-equation imputation
methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by
iteratively fitting a predictive model of one missing attribute, conditional on
all complete attributes and all other missing attributes. Chained-equation
decision trees algorithms are found to be effective~\cite{akande2015empirical}
and are widely used in epidemiological domains~\cite{burgette2010multiple}.

These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
multiple imputation could be incorporated as another avenue of assessing imputation confidence.

In the reference implementation, the chained-equation algorithm proceeds by
iteratively fitting decision trees for the target imputation attributes. In
each iteration, the missing values of a separate attribute are replaced with
newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved. (That is, when the imputed
values of one column are not updated by subsequent iterations that update that
column.) For our experiments, we use Weka's \texttt{REPTree}
implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

@@

\todo{Fill this in.}
\[P(q) = \frac{1}{n} (1 + \overline{Variance(q)}) \]

@@

\todo{Fill this in.}

>>>>

\subsection{Cost model}
\label{sec:cost-model}
The cost of a plan $q$ is a tuple $\langle \textsc{Penalty}(q), \textsc{Time}(q) \rangle$.
$\textsc{Penalty}(q) \in (0,1]$ is a heuristic that quantifies the amount of information lost by the imputation procedure used.
$\textsc{Time}(q) \in (0, \infty)$ is an estimate of the runtime of a query $q$ derived from table statistics, selectivity estimation of the query predicates, and the complexity of the
impute operation.

@@

This set $Q$ will be the Pareto frontier of $S$~\cite{pareto1964cours}, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.

@@

Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$. In essence, $\alpha$ is a tunable parameter 
that determines whether the optimizer focuses on quality or on performance, as plans that
lose a lot of information through dropping are also the fastest.

@@

\noindent{\textbf{Constructing $P$}}: The form of $P$ is chosen to reflect the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations to distribution statistics such as mean and variance.
Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
The heuristic penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases. Finally, the heuristic penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.

@@

\noindent{\textbf{Constructing $T$}}: As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on
the properties of the underlying algorithm via a function $T$. If we assume
that the \textit{Impute} routine is CPU-bound, then we adjust
\textsc{Time} by some constant factor $f$, thereby scaling the result to be
comparable with IO-bound routines.

The form of $T$ depends on the specific imputation method, but there are various
principles useful in capturing the intuition of a practical choice.
$T$ should \emph{increase}
as the number of tuples and complete and dirty attributes passed to the
imputation function increases and it should be take into account the complexity
of the imputation algorithm. Analysis of the time complexity of different
imputation algorithms, especially with in-database execution, is beyond the
scope of this paper. As with $P$, it is the ordering between queries
generated by $T$ that is more important than the precision of estimates.

@@

\subsection{Imputation strategies and their API}
\label{sec:imputation}

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These can then be queried by the optimizer during planning.  The flexibility of
this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation. To
motivate this API, we describe how an analyst would specify the form of these
functions for a general-purpose imputation algorithm, as well as two simpler
methods.

@@

As a reference implementation,
\ProjectName{} uses a general-purpose imputation strategy based on
chained-equation decision trees.  Chained-equation imputation
methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by
iteratively fitting a predictive model of one missing attribute, conditional on
all complete attributes and all other missing attributes. Chained-equation
decision trees algorithms are found to be effective~\cite{akande2015empirical}
and are widely used in epidemiological domains~\cite{burgette2010multiple}.

% Jose: I think this is a bit out of place. So trimmed for space
%These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
%process, in which multiple distinct copies of the complete data are generated and estimators
%are computed by averaging over the multiple datasets to form a single completed dataset~
%\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
%multiple imputation could be incorporated as another avenue of assessing imputation confidence.

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target imputation attributes. In
each iteration, the missing values of a separate attribute are replaced with
newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

@@

At the other end of the spectrum of complexity, an analyst could use a simple
method that replaces a missing value with the mean over available values
in that attribute, a common choice in exploratory data analysis known
as mean value imputation.

We can take advantage of the histograms maintained by \ProjectName{}
to estimate the mean and construct a non-blocking impute operator.
Applying the heuristics from \ref{sec:quality} and \ref{sec:query-runtime}, we
have

\begin{equation*}
\begin{aligned}
    T(q) &\approx 0 \\
    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
\end{aligned}
\end{equation*}

@@

Hot-deck imputation is a nonparametric method in which missing values are
replaced with randomly selected complete values. This method preserves the
distribution of the complete values. The heuristic $P$ remains the same, 
but we modify $T$ to account for the fact that hot-deck is blocking in our
implementation.

\begin{equation*}
\begin{aligned}
    T(q) &= \textsc{Card}(q) \\
    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
\end{aligned}
\end{equation*}
